---
title: "Differences between community models"
subtitle: "Atlas 1, Atlas 2 and Atlas 3"
author: Wessel Mulder 
date: "`r Sys.Date()`"
output: 
  html_document:
  margin:
---
```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	fig.height = 5,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```
```{r helper-functions,include=FALSE}
### HELPER FUNCTIONS 
abbreviate_genus <- function(x) {
  sapply(x, function(nm) {
    parts <- unlist(strsplit(nm, "_"))  # split by underscore
    paste0(substr(parts[1], 1, 1), ". ", parts[2])
  })
}

merge_by_rownames <- function(x,y) {
  merge <- merge(x,y,by='row.names')
  rownames(merge) <- merge$Row.names
  merge[,!colnames(merge) %in% c('Row.names')]
}
```
```{r model-fit-by-guild, include=FALSE}
library(dplyr)
library(ggplot2)
dir <- file.path('.','tmp_rds','mods-complexity-v2')
dirs <- list.dirs(dir,
                  recursive = F)
dirs <- dirs[grepl('.*_(1|2|3)$',dirs)]
dir <- dirs[1]
outputs <- lapply(dirs,function(dir){
  # read objects 
  MF <- readRDS(file.path(dir,'model-outputs','model-fit.rds'))
  VP <- readRDS(file.path(dir,'model-outputs','VP.rds'))
  m <- readRDS(file.path(dir,'m_object.rds'))
  # get the atlas used
  X <- m$XData
  atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
  # get number of occurrences in that atlas 
  if(!length(atlas)==1){
    stop()
  }
  occs <- data.frame(occs = colSums(m$Y,na.rm=T),
                     atlas = atlas)
  row.names(occs) <- colnames(m$Y)
  # get the trait date 
  Tr <- m$TrData
  #transpose VP
  df <-VP$vals # get vals
  df <- as.data.frame(t(df)) # transpose into dataframe 
  # merge VP with species info 
  Tjur <- data.frame(tjur = MF$TjurR2)
  df_rel <- Tjur$tjur*df
  spNames <- m$spNames
  row.names(Tjur) <- spNames
  TrOccs <- merge_by_rownames(Tr,occs)
  TrOccsTjur <- merge_by_rownames(TrOccs,Tjur)
  TrOccsTjurVP <- merge_by_rownames(TrOccsTjur,df_rel)
  TrOccsTjurVP
})
merged <- do.call(rbind,outputs)
str(merged)
colnames(merged) <- c('Migration','Guild','Occurrences',
                      'Atlas','TjurR2',
                      'Temperature','Precipitation','Landscape heterogeneity',
                      "Random effect - site")

# count species per guild
guild_counts <- merged %>%
  group_by(Guild) %>%
  summarise(n_species = n()/3) %>%
  mutate(Guild_counts = paste0(Guild,' (n=',n_species,')')) %>%
  ungroup()

#str(merged)
merged$Guild_counts <- merged$Guild
merged$Guild_counts <- guild_counts$Guild_counts[match(merged$Guild, guild_counts$Guild)]

vars2inspect <- c('Temperature','Precipitation','Landscape heterogeneity')
obs <- lapply(vars2inspect,function(var2inspect){
v2inspect <- sym(var2inspect)

guild_means <- merged %>%
  group_by(Guild_counts, Atlas) %>%
  summarise(!!v2inspect := mean(!!v2inspect, na.rm = TRUE), .groups = "drop")
colors <- c('darkblue','darkgreen','darkred')

ggplot(merged,
       aes(x = reorder(Guild_counts, !!v2inspect, FUN = mean, na.rm = TRUE),
           y=!!v2inspect,
           col = Atlas, fill = Atlas))+
  # violins for guilds with >1 species
  #geom_violin(drop = FALSE, alpha = 0.6) +
  geom_point(alpha = 0.5,size=0.5)+

  # Points for each species, colored by Atlas
  # Points for mean Tjur² per guild per Atlas
  geom_point(data = guild_means,
             aes(y = !!v2inspect, fill = Atlas,color=Atlas),
             size = 1, shape = 21, stroke = 1) +
  
  # set colors 
  scale_color_manual(values=colors)+
  scale_fill_manual(values=colors)+
  
  #scale axes 
  scale_y_continuous(limits=(c(0,0.175)))+
  
  # adjustements 
  labs(title=var2inspect)+
  xlab(label='Guilds (number of species)')+
  ylab(label='Relative variable importance')+
  
  coord_flip()+
  theme_minimal()+
  theme(
    axis.line = element_blank(),             # remove axis lines
    #panel.grid.major = element_blank(),      # remove major gridlines
    #panel.grid.minor = element_blank()       # remove minor gridlines
  ) 
})
```

```{r var-importance-guilds-species, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
dir <- file.path('.','tmp_rds','mods-complexity-v2')
dirs <- list.dirs(dir,
                  recursive = F)
dirs <- dirs[grepl('.*_(1|2|3)$',dirs)]

outputs <- lapply(dirs,function(dir){
  # read objects 
  MF <- readRDS(file.path(dir,'model-outputs','model-fit.rds'))
  VP <- readRDS(file.path(dir,'model-outputs','VP.rds'))
  m <- readRDS(file.path(dir,'m_object.rds'))
  X <- m$XData
  atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
  
  #transpose VP
  df <-VP$vals # get vals
  df <- as.data.frame(t(df)) # transpose into dataframe 
  
  # merge VP with species info 
  Tjur<- data.frame(tjur = MF$TjurR2)
  df_rel <- Tjur$tjur*df

  spNames <- m$spNames
  row.names(Tjur) <- spNames
  Tr <- m$TrData
  species_fit_char <- merge_by_rownames(Tr,Tjur)
  species_fit_char <- merge_by_rownames(species_fit_char,df_rel)
  species_fit_char$atlas <- atlas
  species_fit_char
})

### HOUSEKEEPING
merged <- do.call(rbind,outputs)
rnames <- rownames(merged)
colnames(merged) <- c('Migration','Guild','TjurR2',
                      'Temperature','Precipitation','Landscape heterogeneity',
                      "Random effect - site", 'Atlas')

# count species per guild and add to data 
guild_counts <- merged %>%
  group_by(Guild) %>%
  summarise(Species_counts = n()/3) %>%
  mutate(Guild_counts = paste0(Guild,' (n=',Species_counts,')')) %>%
  ungroup()

merged <- left_join(merged,guild_counts,by='Guild')
rownames(merged)<-rnames
# rename so number of species is included
merged$Atlas <- as.factor(merged$Atlas)
```

```{r guild/tjur plots, include=FALSE, message=FALSE, warning=FALSE}
### GUILD / TJUR PLOTS 
str(merged)

atlases <- c('1','2','3')
guild_tjur_plots <- lapply(atlases,function(atlas){
if(atlas == '1'){year = '1970s'}
if(atlas == '2'){year = '1990s'}
if(atlas == '3'){year = '2010s'}

merged_sub <- merged[merged$Atlas==atlas,]
ggplot(merged_sub,
       aes(x = reorder(Guild_counts, TjurR2, FUN = mean, na.rm = TRUE),
           y= TjurR2))+
  # violins for guilds with >1 species
  geom_violin(drop = FALSE, alpha = 0.6) +
  # violins for guilds with >1 species
  geom_point(data=merged_sub[merged_sub$Species_counts<2,],
             aes(x=Guild_counts,y=TjurR2))+

  #scale axes 
  scale_y_continuous(limits=(c(0,0.7)))+
  
  # adjustements 
  labs(title=paste0('TjurR2 in Atlas ',atlas,' - ',year)) +
  xlab(label='Guilds (number of species)')+
  ylab(label='TjurR2')+
  
  coord_flip()+
  theme_minimal()
})

```

```{r guild/VP plots, echo=FALSE, message=FALSE, warning=FALSE}
### GUILD / VARIABLE PLOTS 
vars2inspect <- c('Temperature','Precipitation','Landscape heterogeneity')
guild_vp_plots <- lapply(vars2inspect,function(var2inspect){
v2inspect <- sym(var2inspect)

guild_means <- merged %>%
  group_by(Guild_counts, Atlas) %>%
  summarise(!!v2inspect := mean(!!v2inspect, na.rm = TRUE), .groups = "drop")
colors <- c('darkblue','darkgreen','darkred')

ggplot(merged,
       aes(x = reorder(Guild_counts, !!v2inspect, FUN = mean, na.rm = TRUE),
           y=!!v2inspect,
           col = Atlas, fill = Atlas))+
  # violins for guilds with >1 species
  #geom_violin(drop = FALSE, alpha = 0.6) +
  geom_point(alpha = 0.5,size=0.5)+

  # Points for each species, colored by Atlas
  # Points for mean Tjur² per guild per Atlas
  geom_point(data = guild_means,
             aes(y = !!v2inspect, fill = Atlas,color=Atlas),
             size = 1, shape = 21, stroke = 1) +
  
  # set colors 
  scale_color_manual(values=colors)+
  scale_fill_manual(values=colors)+
  
  #scale axes 
  scale_y_continuous(limits=(c(0,0.175)))+
  
  # adjustements 
  labs(title=var2inspect)+
  xlab(label='Guilds (number of species)')+
  ylab(label='Absolute variable importance')+
  
  coord_flip()+
  theme_minimal()+
  theme(
    axis.line = element_blank(),             # remove axis lines
    #panel.grid.major = element_blank(),      # remove major gridlines
    #panel.grid.minor = element_blank()       # remove minor gridlines
  ) 
})

```

# What do the model outputs look like? 
### Model summary 
In the past week I've been looking at outputs from the single atlas models I've been working on. A short summary of the model structure:

- all species with a minimum of 5 occurrences in each atlas are included 
  - this makes the species set in all these models identical (~150 species)
- three environmental variables
  - yearly average temperature 
  - yearly total precipitation 
  - landscape heterogeneity (inverse Simpson index of 8 land-use classes)
- dietary guilds, migratory strategy and taxonomy are included 
- data from three atlases 
  - each atlas is modeled separately 

### Model fit
Looking at some diagnostics (TjurR2, as a measure of variance explained) it seems that the three models fit the data reasonably well. In these models, each species gets their own measure of model fit, and in these following plots I've plotted the distribution of these values for each guild. For guilds with multiple species, these are the violins, for guilds with one species, this is just a point with the value of that species. 

Note that the guild-order changes on the y-axis, for each atlas I rearrange them separately based on which guild is (on average) modeled best. What struck me about these plots below is that no guild is consistently modeled best across all atlases. And I find it pretty mixed how 'constant' a guild is modeled, some are much more variable in the individual species fits while others are pretty consistent. 

```{r, printing-guild-tjur-plots, include=T}
guild_tjur_plots[[1]]
guild_tjur_plots[[2]]
guild_tjur_plots[[3]]
```

### Variance partitioning

After this, I started looking into the amount of variance explained by the different variables. For these, I multiply the relative variance explained by the different variables with the total variance explained (the TjurR2 values above). This is to get a measure of absolute variance explained by each variable (I've shown these plots quite a few times now). Species are represented on the Y-axis, and ordered by alfabet (so consistent order across atlases).

This is quite a striking difference between these three. Atlas 2 and 3 show relatively similar responses to the environment, but in Atlas 1 for some reason this explanatory power of the environmental variables almost completely drop. I'm still not entirely sure what's causing this. I'll explore whether it's something do with the data down below. 

I summarize these results in a different way further down, as it's quite hard to compare at this species level across the atlases. 

```{r atlas/VP plots, include=T}
library(dplyr)
library(ggplot2)
dir <- file.path('.','tmp_rds','mods-complexity-v2')
dirs <- list.dirs(dir,
                  recursive = F)
dirs <- dirs[grepl('.*_(1|2|3)$',dirs)]

# personal VP plot function 
plotVariancePartitioning2 =
  function (hM, VP, cols=NULL, main = "Variance Partitioning", ...)
  {
    ng = dim(VP$vals)[1]
    if(is.null(cols)){
      cols = heat.colors(ng, alpha = 1)
    }
    leg = VP$groupnames
    for (r in seq_len(hM$nr)) {
      leg = c(leg, paste("Random: ", hM$rLNames[r], sep = ""))
    }
    means = round(100 * rowMeans(VP$vals), 1)
    for (i in 1:ng) {
      leg[i] = paste(leg[i], " (mean = ", toString(means[i]),
                     ")", sep = "")
    }
    
    VP$vals <- VP$vals[,rev(colnames(VP$vals))]
    barplot(VP$vals, main = main, xlab= " ", ylab = "",las = 1, horiz=T,
            legend = leg, col = cols,
            #las = 2,
            border = NA,
            space=0.1,
            ann=T,
            xlim=c(0,1),
            yaxt='n',
            args.legend = list(x = 'topright',
                               inset = c(0.02,0.05),
                               bg = 'white'),...)
    #rotate 60 degrees (srt = 60)
  }

par(mar = c(5,10,2,2))


s<-lapply(dirs,function(dir){
  # read objects 
  MF <- readRDS(file.path(dir,'model-outputs','model-fit.rds'))
  VP <- readRDS(file.path(dir,'model-outputs','VP.rds'))
  m <- readRDS(file.path(dir,'m_object.rds'))
  X <- m$XData
  atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
  if(atlas == '1'){year = '1970s'}
  if(atlas == '2'){year = '1990s'}
  if(atlas == '3'){year = '2010s'}
  
  VP_2 <- VP

  if(length(VP$groupnames)==1){
    # set plotting parameters 
    if(VP$groupnames == 'tmean_year'){
      alt_name <- c('Yearly temp.')
      colorway <- c('firebrick3','cornsilk2')
    }else if(VP$groupnames == 'prec_year'){
      alt_name <- c('Yearly prec.')
      colorway <- c('dodgerblue3','cornsilk2')
    }else if(VP$groupnames == 'hh'){
      alt_name <- c('Heterogeneity')
      colorway <- c('goldenrod2','cornsilk2')
    }
  }else if(length(VP$groupnames)==3){
    if(identical(VP$groupnames, c('tmean_year','prec_year','hh'))){
      alt_name <- c('Yearly temp.','Yearly prec.','Heterogeneity')
      colorway <- c('firebrick3','dodgerblue3','goldenrod2','cornsilk2')
    }
    if(nrow(VP$vals)-length(VP$groupnames)==2){
      colorway <- c(colorway,'cornsilk3')
    }
  }
  VP_2$groupnames <- alt_name
  tjurs <- MF$TjurR2
  VP_2$vals <- sweep(VP_2$vals, 2, tjurs, "*")
  plotVariancePartitioning2(m,
                            cols = colorway,
                            VP_2,
                            main = paste0('Total variance explained',' - ',year))

})

```

### Variance partitioning / guild
To look a bit more closely into these differences across atlases, I here separated the influence of the environmental variables, and plot the variance explained by each of those variables across guilds. These plots show the average variance explained across all species in a guild, colored by Atlas. The large dots are the averages in the guild, the small dots are the species values. 

I use these plots to look a bit more into detail into what's going on with the first atlas. 

```{r, printing-guild-vp-plots, include=T}
guild_vp_plots[[1]]
guild_vp_plots[[2]]
guild_vp_plots[[3]]
```

```{r preparing-species-env,include=F}
library(ggplot2)
library(ggExtra)
library(cowplot)
library(gridExtra)
library(grid)
library(dplyr)
library(ggplot2)
dir <- file.path('.','tmp_rds','mods-complexity-v2')
dirs <- list.dirs(dir,
                  recursive = F)
dirs <- dirs[grepl('.*_(1|2|3)$',dirs)]
dir <- dirs[1]
outputs <- lapply(dirs,function(dir){
  # read objects 
  #MF <- readRDS(file.path(dir,'model-outputs','model-fit.rds'))
  #VP <- readRDS(file.path(dir,'model-outputs','VP.rds'))
  m <- readRDS(file.path(dir,'m_object.rds'))
  # get the atlas used
  X <- m$XData
  atlas <- unique(sapply(strsplit(rownames(X),'_'),'[',2))
  # get number of occurrences in that atlas 
  if(!length(atlas)==1){
    stop()
  }
  X$atlas <- atlas
  # get occs 
  Y <- m$Y
  # get design
  design <- m$studyDesign
  coords <- m$ranLevels$site$s
  coords$site <- row.names(coords)

  designcoords <- left_join(design,coords,by='site')
  rownames(designcoords) <- rownames(design)
  designX <- merge_by_rownames(designcoords,X)
  designXY <- merge_by_rownames(designX,Y)
})

merged_env <- do.call(rbind,outputs)
```
```{r preparing-oystercatcher,include=F}
species <- 'Haematopus_ostralegus'
species_eng <- 'Eurasian oystercatcher'
subs <- c('atlas','tmean_year','prec_year','hh',species)
subs <- merged_env[,colnames(merged_env) %in% subs]

colnames(subs) <- c('Temperature','Precipitation','Landscape heterogeneity',
                    'Atlas',
                    'PresenceAbsence')
occs_a1 <- sum(subs$PresenceAbsence[subs$Atlas==1])
occs_a2 <- sum(subs$PresenceAbsence[subs$Atlas==2])
occs_a3 <- sum(subs$PresenceAbsence[subs$Atlas==3])
subs$PresenceAbsence <- as.character(subs$PresenceAbsence)
subs$Atlas <- as.factor(subs$Atlas)

combinations <- list(
  c('Temperature','Precipitation'),
  c('Temperature','Landscape heterogeneity')
  )
combination <- combinations[[1]]
# make some plots 
objects<-lapply(combinations,function(combination){
  grid.newpage()
  print(combination)
  v1 <- sym(combination[1])
  v2 <- sym(combination[2])

cols <- c('darkred','darkgreen','darkblue')
alphas <- c(0.1,1)
pmain <- ggplot(data=subs,
       aes(x=!!v1,y=!!v2,col=Atlas,alpha=PresenceAbsence))+
  geom_point()+
  scale_color_manual(values=cols)+
  scale_alpha_manual(values=alphas)+
  labs(title = paste0(abbreviate_genus(species),
                      ' - ',
                      species_eng),
       subtitle = paste0('Atlas 1: n=',occs_a1,', ',
                         'Atlas 2: n=',occs_a2,', ',
                         'Atlas 3: n=',occs_a3))+
  theme_minimal()

print('making density')
### VERSION WITH COLORS 
# Marginal densities along x axis
xdens <- axis_canvas(pmain, axis = "x") +
  ## plot all, with dotted lines 
    geom_density(data = subs, 
               aes(x = !!v1, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               linewidth = 0.5,
               linetype=3)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+

  ## add presences on top
  geom_density(data = subs[subs$PresenceAbsence==1,], 
               aes(x = !!v1, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               linewidth = 0.5,
               linetype=1)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)

# Marginal densities along x axis
ydens <- axis_canvas(pmain, axis = "y",coord_flip=T) +
  geom_density(data = subs, 
               aes(x = !!v2, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               alpha=1,
               linewidth=0.5,
               linetype = 3)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+

    geom_density(data = subs[subs$PresenceAbsence==1,], 
               aes(x = !!v2, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               alpha=1,
               linewidth=0.5,
               linetype=1)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+
coord_flip()

### GET TOGETHER
print('getting together')
p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top")
p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right")
ggdraw(p2)

### ANNOTATE 
# grid.text(label='Presences in:',x=0.87,y=0.725,just='left',
#           gp=gpar(col='black'))+
# grid.text(label=paste0('Atlas 1 - ',occs_a1),x=0.87,y=0.7,just='left',
#           gp=gpar(col='darkred'))+
# grid.text(label=paste0('Atlas 2 - ',occs_a2),x=0.87,y=0.675,just='left',
#           gp=gpar(col='darkgreen'))+
# grid.text(label=paste0('Atlas 3 - ',occs_a3),x=0.87,y=0.65,just='left',
#           gp=gpar(col='darkblue'))

})
```

The plovers (shorebirds that apply a run-stop-peck strategy for foraging) have much variance explained by temperature. The Eurasian oystercatcher is a species with high variance explained in both Atlas 2 and Atlas 3. Below I've plotted the distributions of temperature and precipitation in the atlas-squares associated with the presences and absences of the species.

The dotted lines on the side shows the full distribution of the environmental variables. The solid lines on the side show the distribuitons of the environmental variables for only the presence points. 

What I'm getting from a plot like this is that while in each atlas there seems to be some climatic signature, they're disjunct from one another (e.g. atlas 1/2 overlap in climatic space, but atlas 3 is in a completely different climatic space.) The 'shapes' of the climatic associations are also fairly consistent across the atlases, in positioned in some sort of 'c'. 

I think this could be due to a spatial signature (e.g. habitat), that happens to have a specific climatic signature, which is being picked up by the model. I think this makes sense, especially from the perspective of these shorebirds and a specific relationship between temperature/precipitation on the coasts.

It doesn't really answer the quesion with what's going on with the first atlas data though. 


```{r plotting-oystercatcher}
objects[[1]]
objects[[2]]
```

To get an idea of the original distributions, I plot these below. The plotting in this format looks a bit janky but it's just to get an idea of the distributions. To double check if there's no mistakes made on my end you can easily check [birdlife Denmark](https://dofbasen.dk/atlas/arter/04500/Strandskade)

```{r preparing-oystercatcher-maps,include=T}
library('RColorBrewer')
species <- 'Haematopus_ostralegus'
species_eng <- 'Eurasian oystercatcher'
subs <- c('atlas','X','Y','tmean_year','prec_year','hh',species)
subs <- merged_env[,colnames(merged_env) %in% subs]

colnames(subs) <- c('X','Y',
                    'Temperature','Precipitation','Landscape heterogeneity',
                    'Atlas',
                    'PresenceAbsence')
occs_a1 <- sum(subs$PresenceAbsence[subs$Atlas==1])
occs_a2 <- sum(subs$PresenceAbsence[subs$Atlas==2])
occs_a3 <- sum(subs$PresenceAbsence[subs$Atlas==3])
subs$PresenceAbsence <- as.character(subs$PresenceAbsence)
subs$Atlas <- as.factor(subs$Atlas)

atlas <- 1
library(fields)
# make some plots 
s<-lapply(c(1,2,3),function(atlas){
  # get palettes 
  pal <- brewer.pal(9,'YlGn')
  pal_raw <- pal
  # subset data 
  atlasdata <- subs[subs$Atlas==atlas,]
  # get year 
  if(atlas == '1'){year = '1970s'}
  if(atlas == '2'){year = '1990s'}
  if(atlas == '3'){year = '2010s'}

      # define a binary palette: 0 = light red, 1 = dark red
    binary_cols <- c("indianred1",pal_raw[9])
    
    # assume og_xy_species[,specie] contains 0/1
    og_cols <- binary_cols[as.numeric(atlasdata$PresenceAbsence) + 1]  # +1 because R indexes start at 1
    
    # plot the points
    plot(Y~X, 
         asp=1,
         data = atlasdata,
         col = og_cols, pch = 19,
         xlab = 'X',
         ylab = 'Y',
         main = paste0(year,' - Presence/Absence'))
    # add a binary legend
    image.plot(legend.only = TRUE,
               zlim = c(0,1),                   # 0/1 scale
               col = binary_cols,                # binary colors
               legend.shrink = 0.5,
               axis.args = list(at = c(0,1), labels = c("Absent","Present")),
               legend.lab = "",
               horizontal = T)

})

```


This whole 'habitat' missing from the equation made me curious what would happen if I looked at the the contribution of different land-use coverage percentages (the ones we ended up excluding) from the original analysis. I haven't added these in the models, but below I plot the marginal distributions of the presence/absence points as I did before. I'll use the Eurasian Oystercatcher again. 

```{r preparing-oystercatcher-ocean,include=F}
input <- '.'
env_vars <- 'all'
library(sf)


X <- read.csv(file.path(input,'data/1_preprocessing/X_environmental/X_Environmental.csv'),row.names=1)
X <- X[sort(row.names(X)),]

# get ocean thresholds
grids_thresholds <- st_read(file.path(input,'data/1_preprocessing/atlas-grids/grids-ocean-thresholds/grids_ocean_thresholds.shp'))
thresholds <- grids_thresholds$kvdrtkd[grids_thresholds$pct_lnd>=25]

X <- X[sub("_[123]$", "", rownames(X)) %in% thresholds,]
# any NAs?
table(is.na(X))
if(!env_vars=='all'){X <- X[,env_vars,drop=F]}
table(is.na(X)) # still som NAs
X <- na.omit(X)
table(is.na(X)) # fixed

# grab tmean_year
sites_actual <- row.names(X)

### OCCURRENCES 
Y <- read.csv(file.path(input,'data/1_preprocessing/Y_occurrences/Y_occurrences.csv'),row.names=1)

# remove sites without data 
Y <- Y[row.names(Y) %in% sites_actual,]

merged_landuse <- merge_by_rownames(X,Y)
merged_landuse$atlas <- sapply(strsplit(row.names(merged_landuse),'_'),function(x) x[2])

species <- 'Haematopus_ostralegus'
species_eng <- 'Eurasian oystercatcher'
subs <- c('atlas','tmean_year','LULC_0','LULC_44',species)
subs <- merged_landuse[,colnames(merged_landuse) %in% subs]


colnames(subs) <- c('Temperature','% coverage ocean','% coverage forest',
                    'PresenceAbsence','Atlas')
occs_a1 <- sum(subs$PresenceAbsence[subs$Atlas==1])
occs_a2 <- sum(subs$PresenceAbsence[subs$Atlas==2])
occs_a3 <- sum(subs$PresenceAbsence[subs$Atlas==3])
subs$PresenceAbsence <- as.character(subs$PresenceAbsence)
subs$Atlas <- as.factor(subs$Atlas)

combinations <- list(
  c('Temperature','% coverage ocean'),
  c('Temperature','% coverage forest')
  )
combination <- combinations[[1]]
# make some plots 
objects<-lapply(combinations,function(combination){
  grid.newpage()
  print(combination)
  v1 <- sym(combination[1])
  v2 <- sym(combination[2])

cols <- c('darkred','darkgreen','darkblue')
alphas <- c(0.1,1)
pmain <- ggplot(data=subs,
       aes(x=!!v1,y=!!v2,col=Atlas,alpha=PresenceAbsence))+
  geom_point()+
  scale_color_manual(values=cols)+
  scale_alpha_manual(values=alphas)+
  labs(title = paste0(abbreviate_genus(species),
                      ' - ',
                      species_eng),
       subtitle = paste0('Atlas 1: n=',occs_a1,', ',
                         'Atlas 2: n=',occs_a2,', ',
                         'Atlas 3: n=',occs_a3))+
  theme_minimal()

print('making density')
### VERSION WITH COLORS 
# Marginal densities along x axis
xdens <- axis_canvas(pmain, axis = "x") +
  ## plot all, with dotted lines 
    geom_density(data = subs, 
               aes(x = !!v1, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               linewidth = 0.5,
               linetype=3)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+

  ## add presences on top
  geom_density(data = subs[subs$PresenceAbsence==1,], 
               aes(x = !!v1, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               linewidth = 0.5,
               linetype=1)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)

# Marginal densities along x axis
ydens <- axis_canvas(pmain, axis = "y",coord_flip=T) +
  geom_density(data = subs, 
               aes(x = !!v2, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               alpha=1,
               linewidth=0.5,
               linetype = 3)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+

    geom_density(data = subs[subs$PresenceAbsence==1,], 
               aes(x = !!v2, 
                   #fill = Atlas,
                   col=Atlas),
               size = 0.2,
               alpha=1,
               linewidth=0.5,
               linetype=1)+
  #scale_fill_manual(values=cols)+
  scale_color_manual(values=cols)+
coord_flip()

### GET TOGETHER
print('getting together')
p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top")
p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right")
ggdraw(p2)

### ANNOTATE 
# grid.text(label='Presences in:',x=0.87,y=0.725,just='left',
#           gp=gpar(col='black'))+
# grid.text(label=paste0('Atlas 1 - ',occs_a1),x=0.87,y=0.7,just='left',
#           gp=gpar(col='darkred'))+
# grid.text(label=paste0('Atlas 2 - ',occs_a2),x=0.87,y=0.675,just='left',
#           gp=gpar(col='darkgreen'))+
# grid.text(label=paste0('Atlas 3 - ',occs_a3),x=0.87,y=0.65,just='left',
#           gp=gpar(col='darkblue'))

})
```

It's pretty hard to see from the points, but the marginal distribution shows the species occurs much less in points with 0% ocean coverage, which makes sense. This is the time I had to prepare this document for this meeting so let's discuss further in real life! 


```{r plotting-oystercatcher-ocean}
objects[[1]]
#objects[[2]]
```

<!-- ```{r preparing-woodpecker,include=F} -->
<!-- input <- '.' -->
<!-- env_vars <- 'all' -->
<!-- library(sf) -->


<!-- X <- read.csv(file.path(input,'data/1_preprocessing/X_environmental/X_Environmental.csv'),row.names=1) -->
<!-- X <- X[sort(row.names(X)),] -->

<!-- # get ocean thresholds -->
<!-- grids_thresholds <- st_read(file.path(input,'data/1_preprocessing/atlas-grids/grids-ocean-thresholds/grids_ocean_thresholds.shp')) -->
<!-- thresholds <- grids_thresholds$kvdrtkd[grids_thresholds$pct_lnd>=25] -->

<!-- X <- X[sub("_[123]$", "", rownames(X)) %in% thresholds,] -->
<!-- # any NAs? -->
<!-- table(is.na(X)) -->
<!-- if(!env_vars=='all'){X <- X[,env_vars,drop=F]} -->
<!-- table(is.na(X)) # still som NAs -->
<!-- X <- na.omit(X) -->
<!-- table(is.na(X)) # fixed -->

<!-- # grab tmean_year -->
<!-- sites_actual <- row.names(X) -->

<!-- ### OCCURRENCES  -->
<!-- Y <- read.csv(file.path(input,'data/1_preprocessing/Y_occurrences/Y_occurrences.csv'),row.names=1) -->

<!-- # remove sites without data  -->
<!-- Y <- Y[row.names(Y) %in% sites_actual,] -->

<!-- merged_landuse <- merge_by_rownames(X,Y) -->
<!-- merged_landuse$atlas <- sapply(strsplit(row.names(merged_landuse),'_'),function(x) x[2]) -->

<!-- species <- 'Picus_viridis' -->
<!-- species_eng <- 'Greenwoodpecker' -->
<!-- subs <- c('atlas','tmean_year','LULC_0','LULC_44',species) -->
<!-- subs <- merged_landuse[,colnames(merged_landuse) %in% subs] -->

<!-- colnames(subs) <- c('Temperature','% coverage ocean','% coverage forest', -->
<!--                     'PresenceAbsence','Atlas') -->
<!-- occs_a1 <- sum(subs$PresenceAbsence[subs$Atlas==1]) -->
<!-- occs_a2 <- sum(subs$PresenceAbsence[subs$Atlas==2]) -->
<!-- occs_a3 <- sum(subs$PresenceAbsence[subs$Atlas==3]) -->
<!-- subs$PresenceAbsence <- as.character(subs$PresenceAbsence) -->
<!-- subs$Atlas <- as.factor(subs$Atlas) -->

<!-- combinations <- list( -->
<!--   c('Temperature','% coverage ocean'), -->
<!--   c('Temperature','% coverage forest') -->
<!--   ) -->
<!-- combination <- combinations[[1]] -->
<!-- # make some plots  -->
<!-- objects<-lapply(combinations,function(combination){ -->
<!--   grid.newpage() -->
<!--   print(combination) -->
<!--   v1 <- sym(combination[1]) -->
<!--   v2 <- sym(combination[2]) -->

<!-- cols <- c('darkred','darkgreen','darkblue') -->
<!-- alphas <- c(0.1,1) -->
<!-- pmain <- ggplot(data=subs, -->
<!--        aes(x=!!v1,y=!!v2,col=Atlas,alpha=PresenceAbsence))+ -->
<!--   geom_point()+ -->
<!--   scale_color_manual(values=cols)+ -->
<!--   scale_alpha_manual(values=alphas)+ -->
<!--   labs(title = paste0(abbreviate_genus(species), -->
<!--                       ' - ', -->
<!--                       species_eng), -->
<!--        subtitle = paste0('Atlas 1: n=',occs_a1,', ', -->
<!--                          'Atlas 2: n=',occs_a2,', ', -->
<!--                          'Atlas 3: n=',occs_a3))+ -->
<!--   theme_minimal() -->

<!-- print('making density') -->
<!-- ### VERSION WITH COLORS  -->
<!-- # Marginal densities along x axis -->
<!-- xdens <- axis_canvas(pmain, axis = "x") + -->
<!--   ## plot all, with dotted lines  -->
<!--     geom_density(data = subs,  -->
<!--                aes(x = !!v1,  -->
<!--                    #fill = Atlas, -->
<!--                    col=Atlas), -->
<!--                size = 0.2, -->
<!--                linewidth = 0.5, -->
<!--                linetype=3)+ -->
<!--   #scale_fill_manual(values=cols)+ -->
<!--   scale_color_manual(values=cols)+ -->

<!--   ## add presences on top -->
<!--   geom_density(data = subs[subs$PresenceAbsence==1,],  -->
<!--                aes(x = !!v1,  -->
<!--                    #fill = Atlas, -->
<!--                    col=Atlas), -->
<!--                size = 0.2, -->
<!--                linewidth = 0.5, -->
<!--                linetype=1)+ -->
<!--   #scale_fill_manual(values=cols)+ -->
<!--   scale_color_manual(values=cols) -->

<!-- # Marginal densities along x axis -->
<!-- ydens <- axis_canvas(pmain, axis = "y",coord_flip=T) + -->
<!--   geom_density(data = subs,  -->
<!--                aes(x = !!v2,  -->
<!--                    #fill = Atlas, -->
<!--                    col=Atlas), -->
<!--                size = 0.2, -->
<!--                alpha=1, -->
<!--                linewidth=0.5, -->
<!--                linetype = 3)+ -->
<!--   #scale_fill_manual(values=cols)+ -->
<!--   scale_color_manual(values=cols)+ -->

<!--     geom_density(data = subs[subs$PresenceAbsence==1,],  -->
<!--                aes(x = !!v2,  -->
<!--                    #fill = Atlas, -->
<!--                    col=Atlas), -->
<!--                size = 0.2, -->
<!--                alpha=1, -->
<!--                linewidth=0.5, -->
<!--                linetype=1)+ -->
<!--   #scale_fill_manual(values=cols)+ -->
<!--   scale_color_manual(values=cols)+ -->
<!-- coord_flip() -->

<!-- ### GET TOGETHER -->
<!-- print('getting together') -->
<!-- p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top") -->
<!-- p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right") -->
<!-- ggdraw(p2) -->

<!-- ### ANNOTATE  -->
<!-- # grid.text(label='Presences in:',x=0.87,y=0.725,just='left', -->
<!-- #           gp=gpar(col='black'))+ -->
<!-- # grid.text(label=paste0('Atlas 1 - ',occs_a1),x=0.87,y=0.7,just='left', -->
<!-- #           gp=gpar(col='darkred'))+ -->
<!-- # grid.text(label=paste0('Atlas 2 - ',occs_a2),x=0.87,y=0.675,just='left', -->
<!-- #           gp=gpar(col='darkgreen'))+ -->
<!-- # grid.text(label=paste0('Atlas 3 - ',occs_a3),x=0.87,y=0.65,just='left', -->
<!-- #           gp=gpar(col='darkblue')) -->

<!-- }) -->
<!-- ``` -->

<!-- It's pretty hard to see from the points, but the marginal distribution shows the species occurs much less in points with 0% ocean coverage, which makes sense  -->
<!-- ```{r plotting-woodpecker} -->
<!-- objects[[1]] -->
<!-- objects[[2]] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- input <- '.' -->
<!-- env_vars <- 'all' -->
<!-- library(sf) -->


<!-- X <- read.csv(file.path(input,'data/1_preprocessing/X_environmental/X_Environmental.csv'),row.names=1) -->
<!-- X <- X[sort(row.names(X)),] -->

<!-- # get ocean thresholds -->
<!-- grids_thresholds <- st_read(file.path(input,'data/1_preprocessing/atlas-grids/grids-ocean-thresholds/grids_ocean_thresholds.shp')) -->
<!-- thresholds <- grids_thresholds$kvdrtkd[grids_thresholds$pct_lnd>=25] -->

<!-- X <- X[sub("_[123]$", "", rownames(X)) %in% thresholds,] -->
<!-- # any NAs? -->
<!-- table(is.na(X)) -->
<!-- if(!env_vars=='all'){X <- X[,env_vars,drop=F]} -->
<!-- table(is.na(X)) # still som NAs -->
<!-- X <- na.omit(X) -->
<!-- table(is.na(X)) # fixed -->

<!-- # grab tmean_year -->
<!-- sites_actual <- row.names(X) -->

<!-- ### OCCURRENCES  -->
<!-- Y <- read.csv(file.path(input,'data/1_preprocessing/Y_occurrences/Y_occurrences.csv'),row.names=1) -->

<!-- # remove sites without data  -->
<!-- Y <- Y[row.names(Y) %in% sites_actual,] -->

<!-- for(number in c('1','2','3')){ -->
<!--   Y_sub <- Y[rownames(Y)[grep(paste0("_",number,"$"), rownames(Y))],,drop=F] -->
<!--   if(any(colSums(Y_sub, na.rm =T)<5)){ -->
<!--     print(paste0('In atlas ',number,' these species: ')) -->
<!--     print(names(which(colSums(Y_sub, na.rm =T)<5))) -->
<!--     tofilter <- names(which(colSums(Y_sub, na.rm =T)<5)) -->
<!--     print('have less than 5 occurrences') -->
<!--     Y <- Y[, !(colnames(Y) %in% tofilter)] -->
<!--     print('and are now filtered ') -->
<!--   } -->
<!-- } -->

<!-- for(number in c('1','2','3')){ -->
<!--   Y_sub <- Y[rownames(Y)[grep(paste0("_",number,"$"), rownames(Y))],,drop=F] -->
<!--   if(any(colSums(Y_sub, na.rm =T)<5)){ -->
<!--     print(paste0('In atlas ',number,' these species: ')) -->
<!--     print(names(which(colSums(Y_sub, na.rm =T)<5))) -->
<!--     tofilter <- names(which(colSums(Y_sub, na.rm =T)<5)) -->
<!--     print('have less than 5 occurrences') -->
<!--     stop('stop') -->
<!--   } -->
<!-- } -->

<!-- Tr <- read.csv(file.path(input,"data/1_preprocessing/Tr_aits/traits-guild_migration.csv"),row.names = 2)[,c(2,3)] -->
<!-- # sort by Y -->
<!-- Tr <- Tr[colnames(Y), , drop = FALSE] -->

<!-- merged_landuse <- merge_by_rownames(X,Y) -->
<!-- merged_landuse$atlas <- sapply(strsplit(row.names(merged_landuse),'_'),function(x) x[2]) -->


<!-- specie <- 'Picus_viridis' -->
<!-- species_eng <- 'Greenwoodpecker' -->
<!-- species <- colnames(Y) -->
<!-- means_species <- do.call(rbind,lapply(species,function(specie){ -->

<!--   subs <- c('atlas','tmean_year','prec_year',specie) -->
<!--   subs <- merged_landuse[,colnames(merged_landuse) %in% subs] -->

<!--   colnames(subs) <- c('Temperature','Precipitation', -->
<!--                       'PresenceAbsence','Atlas') -->

<!--   subs$PresenceAbsence <- as.character(subs$PresenceAbsence) -->
<!--   subs$Atlas <- as.factor(subs$Atlas) -->

<!--   combination <- c('Temperature','Precipitation') -->
<!--   atlases <- c(1,2,3) -->
<!--   library(rethinking) -->
<!--   means_atlases <- do.call(rbind,lapply(atlases,function(atlas){ -->


<!--     means_long <- do.call(rbind, lapply(combination, function(combi) { -->
<!--       subs$Temperature_scaled <- standardize(subs$Temperature) -->
<!--       subs$Precipitation_scaled <- standardize(subs$Precipitation) -->

<!--       if(combi == 'Temperature'){variable <- 'Temperature_scaled'} -->
<!--       if(combi == 'Precipitation'){variable <- 'Precipitation_scaled'} -->


<!--       bg <- mean(subs[subs$Atlas == atlas, variable]) -->
<!--       presence <- mean(subs[subs$Atlas == atlas & subs$PresenceAbsence == 1,variable]) -->

<!--       data.frame( -->
<!--         variable = combi, -->
<!--         type = c("bg", "presence"), -->
<!--         mean_value = c(bg, presence), -->
<!--         atlas = atlas, -->
<!--         species = specie -->
<!--       ) -->
<!--     })) ## variables list  -->
<!--   })) ## atlas list -->
<!-- })) ## species list -->
<!-- str(means_species) -->
<!-- str(means_species) -->
<!-- str(Tr) -->
<!-- Tr$species <- rownames(Tr) -->
<!-- means_species$Guild <- Tr$foraging_guild_consensus[match(means_species$species, Tr$species)] -->

<!-- library(tidyr) -->
<!-- diff_means <- means_species %>% -->
<!--   pivot_wider( -->
<!--     names_from = type, -->
<!--     values_from = mean_value -->
<!--   ) %>% -->
<!--   mutate( -->
<!--     diff = presence - bg -->
<!--   ) -->

<!-- diff_means$atlas <- as.factor(diff_means$atlas) -->
<!-- diff_means$Guild <- as.factor(diff_means$Guild) -->

<!-- guilds <- unique(diff_means$Guild) -->
<!-- guild <- 'Terns' -->
<!-- lapply(guilds,function(guild){ -->
<!--   ggplot(data=diff_means[diff_means$Guild==guild,], -->
<!--        aes(x=atlas, -->
<!--            y=diff))+ -->
<!--   geom_boxplot(notch=T)+ -->
<!--   facet_grid(~variable)+ -->
<!--   geom_hline(yintercept=0) -->
<!-- }) -->



<!-- ``` -->



